### 2023-01-27 paper-thesis/chapters/10-coc.tex
A substitution based approach

In Statix, all this information can be stored in a \emph{scope graph}~\cite{scope_graphs}. It is a graph consisting of nodes for scopes, labeled edges for visibility relations, scoped declarations for a relation, and queries for references. We only use a single type of edge, called \verb|P| (parent) edges. It also only has a single relation, called \verb|name|. This name stores a \verb|NameEntry|, which can be either a \verb|NType|, which stores the type of a name, or a \verb|NSubst|, which stores a name that has been substituted with a value. The Statix definition of these concepts is given below:
\begin{lstlisting}
constructors
    NameType : Expr -> NameEntry
    NameSubst : scope * Expr -> NameEntry
relations
    name : ID -> NameEntry
\end{lstlisting}

@@

\begin{lstlisting}
sPutType  : scope * ID * Expr -> scope
sPutSubst : scope * ID * (scope * Expr) -> scope
sGetName  : scope * ID -> NameEntry
sEmpty    : -> scope
\end{lstlisting}
The \verb|sPutType| and \verb|sPutSubst| predicates generate a new scope given a parent scope and a type or a substitution respectively. To query the scope graph use \verb|sGetName|, which will return a \verb|NameEntry| that the query found. Finally, \verb|sEmpty| returns a fresh empty scope.

@@

We implemented beta reduction using a Krivine abstract machine~\cite{krivine}. The machine can head evaluate lambda expressions with a call-by-name semantics. It works by keeping a stack of all arguments that have not been applied yet. This turned out to be the more natural way of expressing this compared to substitution-based evaluation relation. We originally tried to implement the latter, which works fine for the base language. However, it runs into trouble when implementing inductive data types. An additional benefit is that abstract machines are usually more efficient than substitution-based approaches.

@@

Finally, we define \verb|expectBetaEq|. This rule first beta reduces the heads of both sides, and then compares them. If the head is not the same, the rule fails. Otherwise, it recurses on the sub-expressions. One special case is when comparing two \verb|FnConstruct|s. Here we need to take into account alpha equality: two expressions which only differ in the names that they use should be considered equal. We implement this by substituting in the body of the functions, replacing their argument names with a unique placeholder.

@@

\section{Discussion of a substitution-based approach}
\todo{Do this}

>>>>

In Statix, all this information can be stored in a \emph{scope graph}~\cite{scope_graphs}, as explained in chapter \ref{chap:bg-spoofax}. We only use a single type of edge, called \verb|P| (parent) edges. We also only have a single relation, called \verb|name|. This name stores a \verb|NameEntry|, which can be either a \verb|NType|, which stores the type of a name, or a \verb|NSubst|, which stores a name that has been substituted with a value. The Statix definition of these concepts is given below:
\begin{lstlisting}
constructors
    NameType : Expr -> NameEntry
    NameSubst : scope * Expr -> NameEntry
relations
    name : ID -> NameEntry
\end{lstlisting}

@@

\begin{lstlisting}
sPutType  : scope * ID * Expr -> scope
sPutSubst : scope * ID * (scope * Expr) -> scope
sGetName  : scope * ID -> NameEntry
sEmpty    : -> scope
\end{lstlisting}
The \verb|sPutType| and \verb|sPutSubst| predicates generate a new scope given a parent scope and a type or a substitution respectively. To query the scope graph we use \verb|sGetName|, which will return the closest \verb|NameEntry| with a matching same. Finally, \verb|sEmpty| returns a fresh empty scope.

@@

We implemented beta reduction using a Krivine abstract machine~\cite{krivine}. The machine can head evaluate lambda expressions with a call-by-name semantics. It works by keeping a stack of all arguments that have not been applied yet. This turned out to be the more natural way of expressing this compared to substitution-based evaluation relation, which is an alternative we will discuss in section \ref{sec:coc-subst}.

@@

We need to define \verb|expectBetaEq|, which asserts that two scoped expressions are equal under beta reduction. This rule first beta reduces the heads of both sides, and then compares them. If the head is not the same, the rule fails. Otherwise, it recurses on the sub-expressions. One special case is when comparing two \verb|FnConstruct|s. Here we need to take into account alpha equality: two expressions which only differ in the names that they use should be considered equal. We implement this by substituting in the body of the functions, replacing their argument names with a unique placeholder.

@@

\section{Discussion of a substitution-based approach}
\label{sec:coc-subst}

An alternative for a Krivine machine, which keeps a stack of arguments it has encountered, is a substitution-based relation. This beta-reduces a \verb|FnDestruct| by doing a nested beta-reduction of the function, and substituting into that, as is shown in figure \ref{fig:subst-approach}.

Although these rules look cleaner, they are more complicated to implement, requiring a separate relation to check if \verb|f| is a \verb|FnConstruct| or something else. For the pure calculus of constructions, this still works quite well but when adding inductive datatypes (chapter \ref{ch:datatypes}) this quickly becomes a mess.

An additional benefit of Krivine machines is that the run-time performance of them tends to be a bit higher, though this is only a constant factor and not a time-complexity improvement.

\begin{figure}[h]
	\figuresection[\fbox{$\bhr{\scope{s_1}{e_1}}{}{\scope{s_2}{e_2}}$}]{A substitution-based approach for beta reduction}
	\begin{mathpar}
		\mprset{vskip=0.7ex}

		\inferrule{
			\bhr
			{\scope{s}{f}}{}
			{\scope{s_{f'}}{\co{FnConstruct}(x, \_, b)}}
			\\ \bhr
			{\scope{\co{sPutSubst}(s_{f'}, x, (s, a))}{b}}{}
			{\scope{s_{b'}}{b'}}
		} {
			\bhr
			{\scope{s}{\co{FnDestruct}(f, a)}}{}
			{\scope{s_{b'}}{b'}}
		}

		\inferrule{
			\bhr
			{\scope{s}{f}}{}
			{\scope{s_{f'}}{f'}}
			\\ \nexists x e_1 e_2. f' = \co{FnConstruct}(x, e_1, e_2)
		} {
			\bhr
			{\scope{s}{\co{FnDestruct}(f, a)}}{}
			{\scope{s}{\co{FnDestruct}(f, a)}}{}
		}

		\inferrule{
		} {
			\bhr
			{\scope{s}{\co{FnConstruct}(x, a, b)}}{}
			{\scope{s}{\co{FnConstruct}(x, a, b)}}{}
		}

		\mprset{vskip=0ex}
	\end{mathpar}
	\caption{A substitution-based approach for beta reduction}
	\label{fig:subst-approach}
\end{figure}
